---
title: "Combining datasets with `dplyr`"
output: html_notebook
---

## Outline

- Joining datasets


## Joining datasets

In your data wrangling journey, you will often find yourself wanting to combine
one dataframe with some kind of supplementary or partner dataframe. In our
case, we have the penguins and weather data stored separately, but if we ever
wanted to explore any relationships between them, we'd ideally want them in a
single dataframe. This requires lining up the observations and variables
in the datasets appropriately, something which is accomplished by performing
appropriate **joins**.  
  
The key to joining is to identify the variables by which you want to join the
data. That is, we want to ask the question: which columns in each data are the
ones that link them together? In some cases these may be one-to-one matches
(e.g. ID numbers to IDs numbers), or in other cases there is data at different
levels that need to be lined up.


### Left joins

There are several kinds of join function in `dplyr`, but we'll just focus on
`left_join` and leave you to explore the others for yourself.

Like all the join functions, `left_join` takes three arguments: the two dataframes
you'd like to join, and the name of the column (or columns) by which to join.

```
dplyr::left_join(data_left, data_right, by = <cols_to_join_on>)
```

The way `left_join` works is to match up the columns given in the `by` column
and create a new dataframe by pasting new columns from `data_right` alongside
the columns from `data_left`. As many rows from `data_right` are brought in as
possible. The 'left' in `left_join` indicates that we're keeping everything from
the 'left' dataframe (i.e. the first one) and joining the other dataframe *onto*
the 'left' one.

It's helpful to use small, toy dataframes to explore how joins work (or remind
yourself after a period of time away).

```{r}
df1 <- data.frame(colA = c(1, 2, 3, 4),
                  colB = c(2021, 2022, 2023, 2024),
                  colC = c('a', 'b', 'c', 'd'))

df2 <- data.frame(colA = c(1, 3),
                  colD = c('foo', 'bar'),
                  colE = c('dog', 'cat'))
```

```{r}
df1
```

```{r}
df2
```

Now let's join `df2` onto `df1` by the column `colA`:

```{r}
df1 |>
  dplyr::left_join(df2, by = "colA")
```

Notice:

- All the data from the left-hand dataframe, `df1`, is kept (columns `colA` -- `colC`).

- The data from the right-hand dataframe `df2` has been brought over
  for the rows where the 'by' column `colA` has matching values in the
  left-hand dataframe `df1`.
  
- The rows where there aren't matching values in the 'by' column `colA` have
  missing values for the other right-hand dataframe columns (`colD` and `colE`
  in this case).


### Exercise: left join practice

Suppose now the right hand dataframe is as follows:

```{r}
df2 <- data.frame(colA = c(1, 3, 2, 4, 5),
                  colD = c('foo', 'bar', 'baz', 'qux', 'foo'),
                  colE = c('dog', 'cat', 'mouse', 'rabbit', 'horse'))
```

Can you guess what the output of `dplyr::left_join(df1, df2, by = "colA")` will
be? Check your guess with code below.

```{r}
df1 |>
  dplyr::left_join(df2, by = "colA")
```

Now what if you swap the order of `df1` and `df2`? Guess the output of
`dplyr::left_join(df2, df1, by = "colA")` and check your answer below.

```{r}
df2 |>
  dplyr::left_join(df1, by = "colA")
```


### Optional exercise: repeated keys

The values in the 'by' column(s) are sometimes called _keys_ for the join. The
rules we described in the above example are not the whole story and can be
complicated by the presence of repeated keys.

What do you think happens if there are repeated keys? For example, try to
guess the output of the following code:

```{r}
df1 <- data.frame(colA = c(1, 2, 3, 3),  # repeated key 3
                  colB = c(2021, 2022, 2023, 2024),
                  colC = c('a', 'b', 'c', 'd'))

df2 <- data.frame(colA = c(1, 3),
                  colD = c('foo', 'bar'),
                  colE = c('dog', 'cat'))

df1 |>
  dplyr::left_join(df2, by = "colA")
```

Now what if we swap the roles of `df1` and `df2`?

```{r}
df2 |>
  dplyr::left_join(df1, by = "colA")
```


#### Warning about doubles

It's generally not a good idea to join on a 'by' column (or columns)
that are of type double. This is because matching in the join will be done
by an exact equality test on the doubles, which can create strange
results due to numerical imprecision and be difficult to reproduce. Example: in
the following, two doubles that look distinct are considered equal to
`1 / 3 = 0.333...`, so 'match' on the `1 / 3` entry in data frame `x`.

```{r}
x <- data.frame(colA = c(1 / 3, 2 / 3),
                colB = c(1000, 2000))

y <- data.frame(colA = c(0.33333333333333331, 0.33333333333333334),
                colB = c("foo", "bar"))

dplyr::left_join(x, y, by = "colA")
```


### Exercise: joining penguin and weather data

Recall the penguin data from the `palmerpenguins` package. We're going to join
this with some annual weather data, taken from the Palmer Station in Antarctica
from 1989 - 2019[^1].

```{r}
library(palmerpenguins)  # loads `penguins` data
weather <- readr::read_csv("./data/weather_annual.csv")

penguins
weather
```

Use a join to create a single dataframe that has all the penguin data and
weather data combined. Hint: there are some subtleties to be aware of:

- First think about which column(s) to join on -- a call to `dplyr::rename`
  might be in order!

- We should make sure that the 'by' column(s) to join on are of the same type.
  Examine the kind of data in each dataframe closely and coerce if necessary!

```{r}
weather_cleaned <- weather |>
  dplyr::rename(year = Year) |>
  dplyr::mutate(year = as.integer(year))

penguins |>
  dplyr::left_join(weather_cleaned, by = "year")
```


## Acknowledgement

The material in this notebook is adapted from Eliza Wood's [Tidyverse: Data
wrangling & visualization](https://liza-wood.github.io/tidyverse_intro/) course,
which is licensed under [Creative Commons BY-NC-SA
4.0](https://creativecommons.org/licenses/by-nc-sa/4.0/). This in itself is
based on material from [UC Davis's R-DAVIS
course](https://gge-ucd.github.io/R-DAVIS/index.html), which draws heavily on
[Carpentries](https://datacarpentry.org/R-ecology-lesson/) R lessons.


## Footnotes

[^1]: The dataset was derived from the following dataset: Palmer Station
      Antarctica LTER. 2023. Daily weather averages for Palmer
      Station, Antarctica (1989-2023) ver 9. Environmental Data Initiative.
      https://doi.org/10.6073/pasta/3eefb45dbfb784c3cabe3690ea46fe9e (Accessed
      2024-01-08). See the README in the data folder accompanying these
      notebooks for more details.