---
title: "Introducing the Tidyverse"
output: html_notebook
---

## Outline

-   What is the Tidyverse?
-   Dataframes / tibbles
-   Tidy data format


## What is the Tidyverse?

The Tidyverse is self-described as:

> "...an opinionated collection of R packages designed for data science. All
> packages share an underlying design philosophy, grammar, and data
> structures'."
>
> — <https://www.tidyverse.org/>

When you install the `tidyverse` package, you install a suite of packages. These
include the following; we've marked the ones this course will introduce with a
\->:

-   \-> `readr`: Load 'rectangular' data into an R session (e.g. csv files).
-   \-> `dplyr`: Manipulate data (filtering, computing summaries, etc.)
-   \-> `tidyr`: Reshape data (e.g. into 'tidy' format).
-   \-> `stringr`: Working with strings.
-   \-> `lubridate`: Working with dates
-   `ggplot2`: Visualise data.
-   `tibble`: A modern refresh of the core R dataframe, used throughout
    Tidyverse packages.
-   `forcats`: Tools for working with R 'factors', often used for categorical
    data.
-   `purrr`: Tools for working with R objects in a functional way.

These are designed to help you work with data, from cleaning and manipulation to
plotting and modelling. The benefits of the Tidyverse include:

-   They are increasingly popular with large user bases (good for support /
    advice).
-   The packages are generally very well documented.
-   The packages are designed to work together seamlessly and operate well with
    many other modern R packages.
-   Provide features to help you write expressive code and avoid common pitfalls
    when working with data.
-   Built around a consistent philosophy of how to structure data for analysis
    (the 'tidy' data format).


## Dataframes / tibbles

The key data structure that Tidyverse packages are designed to work with is the
*dataframe*.

Actually, this is not quite the whole story. As you read the documentation for
Tidyverse packages, you'll inevitably come across the term *tibble*. This is
essentially the same thing as a dataframe, although there are some minor
differences. For the purposes of this course, you can consider dataframes and
tibbles to be interchangeable and regard them as 'the same thing'.

To demonstrate this, let's consider the Palmer Station penguins data related to
three species of Antarctic penguins from Horst, Hill, and Gorman[^1]. We'll
work with this dataset throughout this course. The data contains size
measurements for male and female adult foraging Adélie, Chinstrap, and Gentoo
penguins observed on islands in the Palmer Archipelago near Palmer Station,
Antarctica between 2007-2009. Data were collected and made available by
Dr Kristen Gorman and the Palmer Station Long Term Ecological Research (LTER)
Program. You can read more about the package on the
[`palmerpenguins` documentation website](https://allisonhorst.github.io/palmerpenguins/index.html). 

To load the dataset, we just import the `palmerpenguins` package and look at
the `penguins` object.

```{r}

```

If we look closely, we see that the class of `penguins` (i.e. the kind of object
it is) is a tibble, indicated by the `"tbl_df"` in the output below:

```{r}

```


## Tidy data

*Tidy data* is a convention that specifies how to arrange our data into a table.
It states that

-   The columns of the table should correspond to the variables in the data.
-   The rows of the table should correspond to observations (or samples) of the
    variables in the data.

Packages in the Tidyverse are designed to work with tidy data.

You can read more about the tidy data format in [Hadley Wickham's Tidy
Data](https://doi.org/10.18637/jss.v059.i10) paper[^2].


### Tidy data: Example

Suppose we have a pet dog and a pet cat and we want to study the average weight
of them by year, to see if there is some kind of association between the
species and the trend in weight variation over time. We could represent this in
a 'matrix-like' format, with the years corresponding to rows and species
corresponding to columns:

| Year | Dog_weight_kg | Cat_weight_kg |
|:-----|:--------------|:--------------|
| 2021 | 21.4          | 8.7           |
| 2022 | 20.9          | 8.4           |
| 2023 | 21.8          | 8.1           |

: Weights of pets by year (non-tidy format)

However, this is not in a tidy format. The variables we're interested in
studying are the year, animal species and weight, but here the columns
correspond to the weights for each kind of species. Instead, we should have a
separate column for the species:

| Year | Species | Weight_kg |
|:-----|:--------|:----------|
| 2021 | dog     | 21.4      |
| 2021 | cat     | 8.7       |
| 2022 | dog     | 20.9      |
| 2022 | cat     | 8.4       |
| 2023 | dog     | 21.8      |
| 2023 | cat     | 8.1       |

: Weights of pets by year (tidy format)


## Acknowledgement

The material in this notebook is adapted from Eliza Wood's [Tidyverse: Data
wrangling & visualization](https://liza-wood.github.io/tidyverse_intro/) course,
which is licensed under [Creative Commons BY-NC-SA
4.0](https://creativecommons.org/licenses/by-nc-sa/4.0/). This in itself is
based on material from [UC Davis's R-DAVIS
course](https://gge-ucd.github.io/R-DAVIS/index.html), which draws heavily on
[Carpentries](https://datacarpentry.org/R-ecology-lesson/) R lessons.


## Footnotes

[^1]: Horst AM, Hill AP, Gorman KB (2020). palmerpenguins: Palmer Archipelago
      (Antarctica) penguin data. R package version 0.1.1.
      <https://allisonhorst.github.io/palmerpenguins/>. doi: 10.5281/zenodo.3960218.

[^2]: Wickham, H. (2014). Tidy Data. *Journal of Statistical Software*,
    *59*(10), 1–23. <https://doi.org/10.18637/jss.v059.i10>
