---
title: "Loading data into R with `readr`"
output: html_notebook
---

## Outline

- Loading data from a csv file
- Other file formats (Excel workbook, Google Sheets)


## Introduction

The package `readr` provides functions for reading data from 'flat' files,
such as csv or other delimited files. The functionality provided by `readr`
offers some advantages over the base R functions available for performing the
same task.


## Loading data from a csv file

You may have used the R function `read.csv` to load data from a csv file into
a dataframe:

```{r}
# Read in ./data/timeseries_data.csv

```

This is fine in a lot of cases, however, note some assumptions have been made.
For example, in the above, the `Id` column is read as an integer and the `Date`
column is read in as a string. In fact, the original data file records the Id
with padded zeros, like so:

```
Id,Date,X_variable,Y_variable
001,2024-02-01,1.1967837194583961,7.900464364159182
002,2024-02-02,-0.7507563344053951,12.690405000912477
003,2024-02-03,-0.7850492689583279,10.280088829160382
004,2024-02-04,-0.6185457481359723,10.073199260115578
005,2024-02-05,0.6188251316696602,9.183892504016947
006,2024-02-06,-1.295559729374875,4.830591718912964
007,2024-02-07,-1.0478610080617479,11.950495760617637
008,2024-02-08,1.2772821688688454,7.008456322119265
009,2024-02-09,-0.06582643175389392,3.4034822391265145
010,2024-02-10,-0.07455295463910701,11.6370557133845
```

What we really want is to read the Id as a string and the Date as a date
object.

The `read_csv` function from the `readr` package is more 'shouty' about
assumptions made when parsing the data, and can have better default
interpretations. For example:

```{r}

```

Notice how it has explicitly told us that

- `Id` is parsed as a string (which is what we wanted!)
- `Date` is parsed as a date (more about dates later in the course).
- `X_variable` and `Y_variable` are parsed as doubles, as expected.

In general, it's good practice to be explicit about how we parse data from
files, rather than assuming the defaults will work. For example, we've no way of
knowing whether the `readr` package maintainers will decide in the future to
interpret padded numbers as in the `Id` column as integers.

We can be explicit about how columns are parsed by using the `col_types`
optional argument. The general way to do this is to supply a list with names
being the column names and values being a _column specifier_,
`readr::col_<data-type>()`:

```
readr::read_csv("path/to/data.csv",
                col_types = list(colA = readr::col_logical(),    # column colA is a boolean
                                 colB = readr::col_integer(),    # column colB is an integer
                                 colC = readr::col_double(),     # column colC is a floating point number
                                 colD = readr::col_character(),  # column colD is a string
                                 colE = readr::col_date(),       # column colE is a date
                                 .default = readr::col_double()  # columns are floating point numbers
                                                                 # unless specified otherwise
                                 ))
```

**Note:** For a full list of the available column specifiers, run
`vignette("readr")` in the R console.

So, to specify that we want to read the `Id` column as a string, the `Date`
column as a date, and the remaining columns as floating point numbers:

```{r}

```


Notice how the informational message about parsing has gone away, because we
explicitly gave the types.


### Exercise: parsing data from files

Change the above so that the `Id` column is parsed as an integer and the
`Date` column is parsed as a string.

```{r}

```


## Other file formats

There are related packages in the Tidyverse ecosystem that provide support for
reading data from file types other than csv:

- `readxl`: package for reading data from Excel workbooks.
- `googlesheets4`: package for reading data from Google Sheets spreadsheets.

We also mention here the `ncdf4` package for reading NetCDF files (not part of
the Tidyverse).
